A forest is more than its trees: haplotypes and inferred ARGs

### 2024-09-30 paper/writeup.tex
added figures, new alg

@@

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying paths in pairs of consecutive tree sequences
    and merge them, given the correct condition.
    For example, if neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree
     \ref{fig:extending_diagram}.
     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
     and the subforest of edges in $T_{k+1}$ which were added to construct 
     $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
     and denoted $F_O$ and $F_O$. \comment{Need better notation. Currently cant think of anything better.}
     For each edge $e \in F_O$ with child $c$,
     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
     If there exists an edge $e'\in F_I$ also with child $c$,
     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
     If there is a common node in both paths excluding $c$, ie. $(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests. The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node. We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $F_I$, see Figure \ref{complicated extending diagram}. We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
%    We do so by identifying short paths (2 edges) in a tree.
%    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
%    If neighboring trees contain edges with 
%    identical ancestor (parent) and descendant (child) nodes, 
%    we assume that there is implied coalescence of haplotypes between nodes, 
%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
%    suppose there exists a path which contains a node $3$,
%    with parent and child nodes $4$ and $0$, respectively.
%    If there is an edge between the nodes $4$ and $0$ in trees 
%    $T_{k+1}$ (or $T_{k-1}$),
%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
%    This action reduces the length of $4 \to 0$ on the genome,
%    and in some cases, completely removes the edge. 
%    We now perform this action on all such ancestral paths 
%    over the entire tree sequence.
%    With algorithm \ref{alg:edge} we only extend existing edges
%    \comment{Note possible change pending algorithm change}
%    over a larger interval on the genome,
%    and remove unnecessary edges in the process.

@@

	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
	\SetKwFunction{ex}{Extend Paths}
	\SetKwFunction{fp}{Find Paths}
	\DontPrintSemicolon
	\LinesNotNumbered
	%\SetAlgoNoEnd
	\input{A tree sequence $\T$}
	\output{New tree sequence,$\T'$}
	\BlankLine
	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
	{
		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
		{\BlankLine
			$\T = \left(T_{|\T|},..., T_1\right)$\;
		}
	}
		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
		\For{edges $e$ in $F_O$}{
			Let $c$ be child node.\;
			$p_c^{F_O} = [c]$\;
			$p_c^{F_I} = [c]$\;
			$\bar{p}_c = []$ Path to be inserted\;
			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
				$p = \tn{parent}(f)$\;
				Append $p$ to $p_c^{F_O}$\;
				$c = p$\;
			}
			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
				$p =\tn{parent}(f)$\;
				Append $p$ to $p_c^{F_I}$\;
				$c = p$\;
			}
			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
				Merge paths up to intersection point\;
				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
				and sort by node time.\;
			}
			Insert $\bar{p}_c$ into $T_{k+1}$\;		
		}
	}\;
	Update $\T$.\;
	\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$, iter=100)}{
	\For{$*$ \KwTo \Range(iter)}{
		$\T' \gets \T$.\fp(fowards=\tn{True})\;
		$\T' \gets \T$.\fp(fowards=\tn{False})\;
		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
			\Break\BlankLine}
	}
	\KwRet{$\T'$}\;
}
\end{algorithm2e}

@@

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7.pdf}
		\caption{Left: Discrepancy \newline Right: True Proportion }
	\end{center}
	\label{fig:discrepancy-over-sample}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s.pdf}
		\caption{Left: Discrepancy \newline Right: True Proportion}
	\end{center}
\end{figure}

@@

@@

\begin{figure}
	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
    \caption{
        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
        as a function of number of trees (modulated by sequence length),
        on reasonably big simulated sequences.
        For "speed": plot ratio of runtime before/after extending.
        \label{fig:speed_and_edges}
    }
\end{figure}

Simple experiment:
(1) simulate, simplify, and extend;
(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".

With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)

@@

@@

\appendix

\begin{figure}
    \caption{
        Supplementary figure: runtime of discrepancy function as a function of number of trees
        (which we modulate by changing sequence length).
        \label{fig:speed_discrepancy}
    }
\end{figure}

>>>>

%\subsection{Extend Edges}
%	\comment{Should we keep a brief section about the original extend edges? If yes, some notes are below.}
%    We do so by identifying short paths (2 edges) in a tree.
%    If neighboring trees contain edges with 
%    identical ancestor (parent) and descendant (child) nodes, 
%    we assume that there is implied coalescence of haplotypes between nodes, 
%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
%    suppose there exists a path which contains a node $3$,
%    with parent and child nodes $4$ and $0$, respectively.
%    If there is an edge between the nodes $4$ and $0$ in trees 
%    $T_{k+1}$ (or $T_{k-1}$),
%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
%    This action reduces the length of $4 \to 0$ on the genome,
%    and in some cases, completely removes the edge. 
%    We now perform this action on all such ancestral paths 
%    over the entire tree sequence.
%    With algorithm \ref{alg:edge} we only extend existing edges
%    \comment{Note possible change pending algorithm change}
%    over a larger interval on the genome,
%    and remove unnecessary edges in the process.

@@

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ of $|\T|$ trees with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$.
     Each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying paths in pairs of consecutive tree sequences
    and merging them, given the correct condition.
    For example, if there exists paths neighboring trees which contain 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree
     \ref{fig:extending_diagram}.
     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
     and the subforest of edges in $T_{k+1}$ which were added to construct 
     $T_{k+1}$ from $T_k$. We call these subforests the out-forest and in-forest, respectively,
     and denoted $F_O$ and $F_O$.
     For each edge $e \in F_O$ with child $c$,
     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
     If there exists an edge $e'\in F_I$ also with child $c$,
     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
     If there is a common node in both paths excluding $c$, ie. $(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests. The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node. We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original in-forest $F_I$, see Figure \ref{complicated extending diagram}. We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.

@@

	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
	\SetKwFunction{ex}{Extend Paths}
	\SetKwFunction{fp}{Find Paths}
	\DontPrintSemicolon
	\LinesNotNumbered
	%\SetAlgoNoEnd
	\input{A tree sequence $\T$}
	\output{New tree sequence,$\T'$}
	\BlankLine
	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
	{
		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
		{\BlankLine
			$\T = \left(T_{|\T|},..., T_1\right)$\;
		}
	}
		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
		\For{edges $e$ in $F_O$}{
			Let $c$ be child node.\;
			$p_c^{F_O} = [c]$\;
			$p_c^{F_I} = [c]$\;
			$\bar{p}_c = []$ Path to be inserted\;
			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
				$p = \tn{parent}(f)$\;
				Append $p$ to $p_c^{F_O}$\;
				$c = p$\;
			}
			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
				$p =\tn{parent}(f)$\;
				Append $p$ to $p_c^{F_I}$\;
				$c = p$\;
			}
			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
				Merge paths up to intersection point\;
				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
				and sort by node time.\;
			}
			Insert $\bar{p}_c$ into $T_{k+1}$\;		
		}
	}\;
	Update $\T$.\;
	\KwRet{$\T$}\;
}
\BlankLine						
\Fn{\ex ($\T$)}{
	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;

	}
	\KwRet{$\T'$}\;
}
\end{algorithm2e}

@@

 We note that the discrepancy function is a metric for detecting non-similarity between two tree sequences. However, we note that this function is not a metric in the mathematical sense, as it is not symmetric, nor is it reflexive. For example, given a tree sequence $\T$ and its simplification $\mathbb{S}$ and $\T\neq\mathbb{S}$ we have $\operatorname{discrepancy}(\mathbb{S}, \T)=0$. Additionally, given Figure \ref{fig:conceptual_discrepancy}, $\operatorname{discrepancy}(T1, T2) \neq \operatorname{discrepancy}(T2, T1)$ \comment{actually compute this. We may just do it in this section instead of the appendix?}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		\caption{Left: Discrepancy \newline Right: True Proportion }
	\end{center}
	\label{fig:discrepancy-over-sample}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
		\caption{Left: Discrepancy \newline Right: True Proportion}
	\end{center}
	\label{fig:discrepancy-over-seqlen}
\end{figure}

@@

When observing our algorithm, we have to observe that for general tree sequences that have been initially simplified, there are portions of the tree sequence that the extend haplotypes method will not be able to recover. We call these portions the trapped unary spans. Most of these spans occur in the oldest portions of the ancestry near roots of the tree sequence. Thus we must disregard these unary spans when observings the effects of extend haplotypes.

For a tree sequence $\T$ simulated in `msprime` of length $5\times 10^7$ with 1000 samples, recombination rate $10^{-8}$ and population size $10^4$, we found that there

@@

%\begin{figure}
%	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
%    \caption{
%        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
%        as a function of number of trees (modulated by sequence length),
%        on reasonably big simulated sequences.
%        For "speed": plot ratio of runtime before/after extending.
%        \label{fig:speed_and_edges}
%    }
%\end{figure}

%Simple experiment:
%(1) simulate, simplify, and extend;
%(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
%then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
%
%With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
%(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
	\caption{Observation of Extendhapltypes with gene conversion = some number
	Left: 
	Middle:
	Right: }
\end{figure}

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
	\caption{Observation of Extendhapltypes with recombination = some number
		Left: 
		Middle:
		Right: }
\end{figure}

@@

\appendix
%
%\begin{figure}
%    \caption{
%        Supplementary figure: runtime of discrepancy function as a function of number of trees
%        (which we modulate by changing sequence length).
%        \label{fig:speed_discrepancy}
%    }
%\end{figure}
